---
title: "Automate on-brand, accessible data visualisations for your reporting using R and `doestyle`"
author: "Sam Gardiner (and you!)"
date: "4 June 2026"
format:
  live-html:
    fig-cap-location: margin
    reference-location: margin
    citation-location: margin
    toc: true
bibliography: references.bib
engine: knitr
knitr:
  opts_chunk: 
    dev: svglite
webr:
  packages:
    - doestyle
  repos:
    - https://nsw-education.r-universe.dev/
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

# Introduction 

This workshop has been developed and delivered on [Dharug](https://www.cityofparramatta.nsw.gov.au/community/first-nations/dharug-and-first-nations-community) country.

## Abstract

A picture tells a thousand words - but it can be time-consuming to craft data visualisations that meet our accessibility obligations while remaining on-brand and visually appealing. This workshop for R users will guide participants in the construction of several data visualisations using `doestyle`, the department’s in-house R package for plot styling. We will explore workflows to improve the ease and speed at which data visualisations can be built into evaluation reporting, and how to tweak these workflow defaults to improve visual appeal.

## Learning intentions

- Use R, `ggplot2` and `doestyle` to apply styling and theming to data visualisations
- Understand accessibility requirements for colour and contrast, and ensure that your data visualisations are accessible to the broadest section of our community
- Develop workflows to make building multiple plots for reporting as quick and simple as possible

# Getting started

This workshop is designed to be completed in a web browser. It uses [WebR](https://docs.r-wasm.org/webr/latest/) and [Quarto Live](https://r-wasm.github.io/quarto-live/) to run a [WebAssembly](https://webassembly.org/)-powered R session within your browser, with no installation required.

:::{.callout-note}
Whenever you see a 'Run Code' button, you'll need to click it to progress. You can also run the code in a block with Ctrl+Enter (Windows/Linux) or Command+Enter (Mac), if your text cursor is within the block. You can activate auto-completion with Ctrl+Space. You can edit any of the blocks and run them to see how your changes affect the results. Some of the blocks contain exercises - usually you'll need to replace sections of code that are missing, denoted with `______`.
:::


We'll be using a few packages throughout the rest of the workshop, so load them now:

```{r}
#| label: load-packages
#| exercise: load-packages
#| autorun: true
#| output: false
library(doestyle)
library(dplyr)
library(ggplot2)
library(forcats)
```

```{r}
#| label: overwrite-theme-doe
#| aurotrun: true
#| echo: false
#| output: false

# We need to modify theme_doe() slightly for webr. By default it uses the Public
# Sans face, which in turn loads `showtext`, which breaks text rendering in
# webr. We'll wrap it to change the default to sans, instead.
theme_doe <- function(base_size = 16, base_family = "sans", ...) {
  doestyle::theme_doe(base_size, base_family, ...)
}
```

As you loaded this web page, the `doestyle` R package was installed for your use during the workshop. However, if you want to use `doestyle` for your own work in RStudio or Visual Studio Code, you'll need to install it on your own computer (but not today):

```r
install.packages('doestyle', repos = c('https://nsw-education.r-universe.dev',
                                       'https://cloud.r-project.org'))
```

The `doestyle` package includes a dataset called `public_schools`, which is a snapshot of the [NSW public schools master dataset](https://www.metadata.nsw.gov.au/item/9544/distribution/nsw-public-schools-master-dataset).

Let's check that everything is working by building a quick visualisation with `ggplot2`, `doestyle` and the public schools data. Run the code and have a look at the output:

```{r}
#| label: fig-check

ggplot(public_schools,
       aes(y = Operational_directorate)) +
  geom_bar() +
  theme_doe() 
```

You might notice that we didn't have to specify axis labels. This is because newer versions of `ggplot2` automatically use column label metadata, if it exists [@ggplot2400-labels]. The `public_schools` object has column labels that were extracted from our metadata repository at [Metadata.NSW](https://www.metadata.nsw.gov.au/item/9544/distribution/nsw-public-schools-master-dataset).

# A quick tour of `ggplot2`, as helped by `doestyle`

## The Grammar of Graphics + `doestyle`

- `ggplot2` lets you map your data onto geometries via aesthetics and scales, building up plots in layers
- `doestyle` provides helpers for `ggplot2`:
  - Department of Education-branded scales
  - easy access to NSW Government-branded colours
  - a couple of custom geometries
  - tools to check colour contrast ratios and relative luminosity
  - tools to make it slightly easier to use the Public Sans typeface

If you've never used `ggplot2` before, there are plenty of good resources to get started:

- Hadley Wickham's paper introducing the package and its implementation of the grammar of graphics [@wickhamGgplot22011]
- The amazing, free and frequently-updated offical ggplot2 book [@wickhamGgplot2ElegantGraphics2026]
- Jonathan's alliterative conference talk on the grammar of graphics [@mcguireCharmingChartsGentle2024]

## Warm-up exercises: colours, scales and themes

We'll use the copy of `public_schools` included within `doestyle`, and filter it so we're just looking at primary and secondary schools that have enrolments (`latest_year_enrolment_FTE`), a reported LBOTE percentage (`LBOTE_pct`) and a reported ICSEA (`ICSEA_value`).

```{r}
#| label: dataset-enrolled
enrolled <- public_schools |>
  # Keep primary and secondary schools
  filter(Level_of_schooling %in% c("Primary School",
                                   "Secondary School")) |>
  # Drop the schools without enrolments, LBOTE proportion or ICSEA
  filter_out(
    when_any(
      is.na(latest_year_enrolment_FTE),
      is.na(LBOTE_pct),
      is.na(ICSEA_value)
    )
  ) |>
  # Keep just the columns we're interested in visualising
  select(School_code,
         School_name,
         Operational_directorate,
         latest_year_enrolment_FTE,
         Level_of_schooling,
         LBOTE_pct,
         ICSEA_value,
         Date_1st_teacher)

# Take a peek:
glimpse(enrolled)
```

Let's visualise the relationship between school size and LBOTE percentage:

```{r}
#| label: size-vs-lbote1
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma())
```

We've made the following declarations, and `ggplot2` has taken care of the rest:

- Map enrolment count values onto the x-axis
- Map LBOTE proportion values onto the y-axis
- Map level of schooling values onto colour
- Display each of these mappings with the 'point' geometry
- Tweak the x-axis scale so that number format includes a comma

The plot works and tells a story, but it doesn't use the NSW Education brand palette. You can look up our colours in our brand guidelines, but the point of `doestyle` is to reduce the friction involved in making on-brand graphics.

```{r}
#| autorun: true
doestyle::show_colours()
```

If we need them, we can also use the whole NSW brand palette:

```{r}
#| column: page
#| fig-width: 10
#| fig-height: 4
#| autorun: true
show_colours("nsw")
```

We can also get the hex code for any named brand colours, without having to look at the swatches:

```{r}
doe_colours("blue-01")
```

Try it for yourself. Pick any two DoE or NSW colour names and retrieve their hex values:

```{r}
#| exercise: pick_colours

doe_colours("______", "______")
```


Now, use those hex codes in the plot we built earlier. We've added on `scale_colour_manual()`, which lets you modify the colour mapping from the `Level_of_schooling` data to the aesthetics used by `geom_point`. Add the hex codes your retreived:

```{r}
#| label: exercise-use-colours-hex
#| exercise: use-colours-hex
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_colour_manual(values = c("______", "______"))
```

Copying and pasting hex codes isn't a great workflow, so let's just use `doe_colours()` inside `scale_colour_manual()`:

```{r}
#| label: exercise-use-scale-colour-manual
#| exercise: use-scale-colour-manual
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_colour_manual(values = doe_colours("______", "______"))
```

So that you don't have to look up colours for every plot, `doestyle` includes a ready-made scale and palettes that you can use instead of `ggplot2::scale_colour_manual()`: `scale_colour_doe()`.

```{r}
#| label: demo-scale-colour-doe
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_colour_doe()
```

Our `scale_colour_doe()` function has a `palette` argument. Its default is `doe_palettes$default`, which provides on-brand defaults for up to 12 categories^[Not that you should ever be trying to produce a plot that uses 12 colours - if so, it's time to rethink what you're trying to visualise!]:

```{r}
#| label: fig-demo-palette
#| fig-height: 1
#| fig-width: 4
plot(doe_palettes$default)
```

Other built-in palettes are available in the `doe_palettes` object, which you can explore on your own.

If you want to define your own palettes, you can:

```{r}
#| label: demo-custom-palette
my_custom_palette <- doe_colours("grey-01", "blue-02")

ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_colour_doe(palette = my_custom_palette) 
```

This will come in handy if you're producing a series of plots with similar categories. Having the same colour language across your plots will reduce the cognitive burden of interpreting them.

Try it yourself. Create a palette with two on-brand colours of your choice, and apply the palette to the plot with `scale-colour_doe()`:

```{r}
#| label: exercise-custom-palette
#| exercise: custom-palette
______ <- doe_colours("______", "______")

ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_colour_doe(palette = ______) 
```

Finally, if you don't want to worry about setting scales or designing a palette, you can just apply `theme_doe()`, which will do its best to use sensible^[Sensible for a couple of colours, but after that you are better off doing your own decision-making.] defaults. You'll notice that `theme_doe()` applies some other defaults, like removing the panel background, increasing most text sizes and reducing the number of gridlines:

```{r}
#| label: demo-theme-doe
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  theme_doe()
```

Depending on the medium for which you're designing your plots, you may want to tweak the `theme_doe()` defaults. If you are using more than one geom (for example, a labelled bar plot - `geom_col()` and `geom_text()`) you will definitely need to start thinking about which colour and fill scales to use.

There's still plenty more we could do to polish this plot, but for now we'll move on to our main topic.

# Accessibility

From our very own website [@doeAccessibilityGuidelines2026]:

> Accessibility means providing equal access to information for all our audiences, with special consideration for people with disability.
> 
> ...
>
> Our approach to accessibility at the NSW Department of Education is about creating content that does not exclude anyone because of their ability, situation or circumstance. Rather, it is an inclusive approach to content that should make it easier for all users to understand, interact with and respond to our websites, apps, communications and materials.

- Accessibility improves the experience for everyone, not just people with a disability
- Starting your visualisations with accessibility in mind will often help you to design a simpler, more interpretable plot

## The standard: WCAG 2.2

We aim for level AA conformance with Web Content Accessibility Guidelines 2.2 [@wcag22].

> Web Content Accessibility Guidelines (WCAG) explains how to make web content more accessible to people with disabilities. WCAG covers web sites, applications, and other digital content.
>
> There are three levels of conformance:
>
> - Level A is the minimum level.
> - Level AA includes all Level A and AA requirements. Many organizations strive to meet Level AA.
> - Level AAA includes all Level A, AA, and AAA requirements. 

[@w3cwebaccessibilityinitiativewaiWebContentAccessibility]

You can read [the whole guideline](https://www.w3.org/TR/WCAG22/), but for our purpose today, we're most interested in sections of [Guideline 1.4 Distinguishable](https://www.w3.org/TR/WCAG22/#distinguishable):

> Make it easier for users to see and hear content including separating foreground from background.

### Colour

> #### [Success Criterion 1.4.1 Use of Color](https://www.w3.org/TR/WCAG22/#use-of-color)
>
> Color is not used as the only visual means of conveying information ...

:::{.callout-caution}
About 1 in 20 of your audience will have some form of colour blindness. If you use colour as the only way of communicating, you risk misinterpretation or even complete failure of your visusalisation.
:::

### Contrast

> #### [Success Criterion 1.4.3 Contrast (Minimum)](https://www.w3.org/TR/WCAG22/#contrast-minimum)
> 
> The visual presentation of text and images of text has a contrast ratio of at least 4.5:1 ...
> 
> Large-scale text and images of large-scale text have a contrast ratio of at least 3:1.

:::{.aside}
_Large text_ means 18 points, or 14 points and bold.
:::

The WWC provides excellent guidance and examples [@worldwidewebconsortiumWebContentAccessibility2024].

- [Quick reference with examples](https://www.w3.org/WAI/WCAG22/quickref/#contrast-minimum)

## How can we meet the standard?

If we use colour to convery information, often we'll need to add a second aesthetic mapping (in addition to colour) to represent categories. Depending on your plot type and data, there are many workable options:

- Colour + shape
- Colour + pattern or line type
- Colour + text annotation

If you have more than about 3 categories, there's a good chance you'll be better off designing a simpler visual:

- Can you move your coloured categories to a labelled axis?
- Can you use small multiples instead (e.g. `ggplot2::facet_wrap()` or `facet_grid()`)?
- Would a table be easier to interpret?

## Getting away from colour alone

For this exercise, we'll look at schools with an attached preschool. Here's our dataset:

```{r}
#| label: preschools-setup
preschools <- public_schools |>
  filter(Preschool_ind == "Y") |>
  filter_out(when_any(is.na(latest_year_enrolment_FTE),
                      is.na(ICSEA_value))) |>
  select(School_code,
         Level_of_schooling,
         Operational_directorate,
         latest_year_enrolment_FTE,
         ICSEA_value)

# Take a peek:
glimpse(preschools)
```

We're counting the number of schools with an attached preschool, by directorate. Here's a plot that uses colour alone to distinguish between the types of school:

```{r}
#| label: preschools-base
ggplot(preschools,
       aes(y = Operational_directorate,
           fill = Level_of_schooling)) +
  geom_bar(position = position_dodge(preserve = "single"),
           width = 2/3) +
  labs(x = "Attached preschools") +
  theme_doe(base_size = 12)
```

We can improve the accessibility of the plot by using `linetype` in addition to colour:

```{r}
#| label: preschool-with-linetype
ggplot(preschools,
       aes(y = Operational_directorate,
           fill = Level_of_schooling,
           # Map level of schooling to a second aesthetic, linetype:
           linetype = Level_of_schooling)) +
  geom_bar(position = position_dodge(preserve = "single"),
           width = 2/3) +
  coord_cartesian(expand = c(left = FALSE)) +
  theme_doe(base_size = 12)
```

We should be able to make the plot a little bit more attractive, while still overloading the colour aesthetic, by choosing colour-linetype combinations more carefully, using `scale_fill_manual()` and `scale_linetype_manual()`.

```{r}
#| label: preschool-with-linetype-prettier
ggplot(preschools,
       aes(y = Operational_directorate,
           fill = Level_of_schooling,
           # Map level of schooling to a second aesthetic, linetype:
           linetype = Level_of_schooling)) +
  geom_bar(position = position_dodge(preserve = "single"),
           width = 2/3,
           # Use the same line colour for every category
           color = doe_colours("blue-01")) +
  scale_fill_manual(values = doe_colours("Primary School" = "blue-01",
                                           "Central/Community School" = "red-03",
                                           "Infants School" = "blue-03")) +
  scale_linetype_manual(values = c("Primary School" = "solid",
                                   "Central/Community School" = "solid",
                                   "Infants School" = "dashed")) +
  coord_cartesian(expand = c(left = FALSE)) +
  theme_doe(base_size = 12)
```

Is it worth trying a different type of plot altogether?

```{r}
#| label: preschools-small-multiples
ggplot(preschools,
       aes(y = Level_of_schooling)) +
  geom_bar() +
  facet_wrap(~Operational_directorate,
             labeller = label_wrap_gen()) +
  coord_cartesian(expand = c(left = FALSE)) +
  theme_doe(base_size = 12, panel.border = element_rect())
```

### Exercise - Getting away from colour alone

It's your turn. Using the same `preschools` dataset, we want to investigate the relationship between school size and its ICSEA.

```{r}
#| label: demo-preschool-colour
ggplot(preschools,
       aes(x = ICSEA_value,
           y = latest_year_enrolment_FTE,
           colour = Level_of_schooling)) +
  geom_point() +
  theme_doe()
```

There's plenty we can learn from this plot, but unfortunately it relies on the colour aesthetic alone for differentiation between school types.

How could we improve it? Make your changes here:

```{r}
#| label: exercise-preschool-colour
#| exercise: preschool-colour
ggplot(preschools,
       aes(x = ICSEA_value,
           y = latest_year_enrolment_FTE,
           colour = Level_of_schooling)) +
  geom_point() +
  theme_doe()
```

:::{.hint exercise="preschool-colour"}
Check out the documentation for [`scale_shape()`](https://ggplot2.tidyverse.org/reference/scale_shape.html#details).
:::

::: {.solution exercise="preschool-colour"}
```{r}
#| label: exericse-preschool-colour-solution
# This is how I tackled the problem - your solution might be better!
ggplot(preschools,
       aes(x = ICSEA_value,
           y = latest_year_enrolment_FTE,
           # Use both colour and shape to distinguish school type
           colour = Level_of_schooling,
           shape = Level_of_schooling)) +
  # Make the points larger
  geom_point(size = 2) +
  # Carefully choose a colour scale
  scale_colour_manual(values = doe_colours("Primary School" = "blue-01",
                                           "Central/Community School" = "red-02",
                                           "Infants School" = "blue-02")) +
  # Carefully choose a shape scale
  scale_shape_manual(values = c("Primary School" = "square",
                                "Central/Community School" = "asterisk",
                                "Infants School" = "plus")) +
  # Format the x-axis so it uses commas
  scale_x_continuous(labels = scales::label_comma()) +
  theme_doe()
```
:::

### Text contrast

Sometimes your stakeholders will ask for data labels or other text annotations^[An appropriate question when asked for data labels is "Would a table be clearer?"]. 

```{r}
#| label: demo-lbote-topn
# This is a bit silly, but we're going to plot the school with the highest
# LBOTE proportion in each directorate:
lbote_topn <- public_schools |>
  filter_out(is.na(LBOTE_pct)) |>
  slice_max(LBOTE_pct, n = 1, by = Operational_directorate, with_ties = FALSE) |>
  mutate(
    school_pct_label = paste0(LBOTE_pct, "% "),
    school_name_fct = forcats::fct_reorder(School_name, LBOTE_pct))

ggplot(lbote_topn,
       aes(x = LBOTE_pct,
           y = school_name_fct)) +
  # Layers:
  ## The bars on which we'll draw text
  geom_col(width = 4/5) +
  ## Left-hand text (school name)
    geom_text(aes(label = School_name),
            x = 1,
            hjust = "left",
            size = 14,
            size.unit = "pt") +
  ## Right-hand text (LBOTE %)
  geom_text(aes(label = school_pct_label),
            hjust = "right",
            size = 14,
            size.unit = "pt") +
  # Add % suffix to x axis
  scale_x_continuous(labels = \(x) paste0(x, "%")) +
  # Add our theme, tweaked to remove the now-superfluous y-axis labels
  theme_doe(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    panel.grid = element_blank())
```

If we want to meet WCAG 2.2 Success Criterion 1.4.3 Contrast (Minimum) (Level AA), we need to make sure that text has a contrast ratio of at least 4.5:1 against its background. For WCAG, W3C defines contrast ratio as 

$$
\text{CR} = \frac{L_1 + 0.05}{L_2 + 0.05}
$$

where $L_1$ and $L_2$ are the relative luminances of the brighter and darker colours being compared. The function for relative luminance is piecewise and depends on the red-, green- and blueness of the colour [@relativeluminance].

Fortunately, `doestyle` provides some help with choosing contrasting colours for text.

- `contrast_ratio()`
- `relative_luminance()`

Contrast ratio has a maximum of 21:

```{r}
#| label: maximum-contrast-ratio

# For example: black text on a white background
contrast_ratio(colour1 = doe_colours("white"), colour2 = doe_colours("black"))
```

... and a minimum of 1:

```{r}
#| label: minimum-contrast-ratio
# For example: black text on a black background
contrast_ratio(doe_colours("black"), doe_colours("black"))
```

The Department of [Education Brand Guidelines](https://brand.education.nsw.gov.au/content/edam/en/brand-guideline.html) provide a helpful matrix of combinations that meet the 4.5:1 contrast requirement, but it is helpful to be able to compute them in R, as well. Suppose you want to check several combinations at once:

```{r}
#| label: check-multiple-ratios
# Check red-01, blue-01, grey-01 and orange-02 against a white background:
contrast_ratio(
  colour1 = doe_colours("red-01", "blue-01", "grey-01", "orange-02"),
  colour2 = doe_colours("white")
)
```

You could even somewhat automate the check with some logic, to make sure the palette you've chosen meets the 3.5:1 ratio. In fact, we do something like this in the automated tests that `doestyle` goes through before a release:

```{r}
#| label: test-multiple-ratios
# Check red-01, blue-01, grey-01 and orange-02 against a white background:
contrast_ratio(
  colour1 = doe_colours("red-01", "blue-01", "grey-01", "orange-02"),
  colour2 = doe_colours("white")
) > 4.5
```

### Exercise: find adequate text contrast

We'll be looking at the proportion of schools with opportunity classes:

```{r}
#| label: dataset-oc
oc <- public_schools |>
  filter_out(Operational_directorate == "Unassigned") |>
  # Pre-summarise our counts, so we can compute proportions
  count(Operational_directorate, Opportunity_class) |>
  mutate(.by = Operational_directorate,
         proportion = n / sum(n),
         pct_label = paste0(round(proportion * 100, 1), "%")
  )

# Take a peek:
glimpse(oc)
```

Can you improve the text contrast of this plot?

```{r}
#| label: oc-before
ggplot(oc, aes(x = proportion,
               y = Operational_directorate,
               fill = Opportunity_class,
               label = pct_label)) +
  # The bars upon which we'll draw text
  geom_col(position = position_stack()) +
  # Text layer: label for opportunity class == "Y"
  geom_text(x = 0.01,
            hjust = 0.01,
            data = ~filter(.x, Opportunity_class == "Y")) +
  # Tweak the appearance of our x and y scales
  scale_x_continuous(name = "Percentage of schools with Opportunity Class",
                     labels = scales::label_percent()) +
  scale_y_discrete(name = NULL,
                   labels = scales::label_wrap(25)) +
  theme_doe() 
```

Step 1: find a colour that has a contrast ratio of at least 4.5:1 when plotted over `blue-01` and `red-02`:

```{r}
#| label: exercise-find-ratio
#| exercise: find-ratio
contrast_ratio(
  colour1 = ______,
  colour2 = doe_colours("blue-01", "red-02")
)
```

Step 2: apply your high-contrast colour to the text layer:

```{r}
#| label: exercise-improve-contrast
#| exercise: improve-contrast
ggplot(oc, aes(x = proportion,
               y = Operational_directorate,
               fill = Opportunity_class,
               label = pct_label)) +
  # The bars upon which we'll draw text
  geom_col(position = position_stack()) +
  # Text layer: label for opportunity class == "Y"
  geom_text(x = 0.01,
            hjust = 0.01,
            data = ~filter(.x, Opportunity_class == "Y")
            colour = ______) +
  # Tweak the appearance of our x and y scales
  scale_x_continuous(name = "Percentage of schools with Opportunity Class",
                     labels = scales::label_percent()) +
  scale_y_discrete(name = NULL,
                   labels = scales::label_wrap(25)) +
  theme_doe() 
```

::: { .solution exercise="improve-contrast"}
```{r}
#| label: solution-improve-contrast
ggplot(oc, aes(x = proportion,
               y = Operational_directorate,
               fill = Opportunity_class,
               label = pct_label)) +
  # The bars upon which we'll draw text
  geom_col(position = position_stack()) +
  # Text layer: label for opportunity class == "Y"
  geom_text(x = 0.01,
            hjust = 0.01,
            data = ~filter(.x, Opportunity_class == "Y"),
            # I went with white, and also made the text a bit bigger
            colour = "white",
            size = 14,
            size.unit = "pt") +
  # Tweak the appearance of our x and y scales
  scale_x_continuous(name = "Percentage of schools",
                     labels = scales::label_percent()) +
  scale_y_discrete(name = NULL,
                   labels = scales::label_wrap(25)) +
  theme_doe() 
```
:::

### `text_colours()` and `secondary_colours()`

`doestyle` provides some pre-baked text and secondary colours so you don't have to go searching for them.

```{r}
#| label: demo-text-colours
#| fig-height: 1
doe_colours("blue-01", "red-02")
text_colours("blue-01", "red-02")

plot(doe_colours("blue-01", "red-02"))
plot(text_colours("blue-01", "red-02"))
```

```{r}
#| label: demo-secondary_colours
#| fig-height: 1
doe_colours("blue-01", "red-02")
secondary_colours("blue-01", "red-02")

plot(doe_colours("blue-01", "red-02"))
plot(secondary_colours("blue-01", "red-02"))
```

If these secondary and text colours look familiar, it's because they're used in `show_colours()`:

```{r}
#| label: demo-show-colors-secondary-text
doestyle::show_colours()
```


```{r}
school_gender <- public_schools |>
  count(Operational_directorate, School_gender) |>
  mutate(.by = Operational_directorate,
         proportion = n / sum(n))

ggplot(school_gender,
       aes(x = proportion,
           y = Operational_directorate,
           fill = School_gender)) +
  geom_col()
```

# Workflows

## Bundling customisations

If you're producing multiple plots with similar subject matter, it's helpful to define common scales. If you store them in a standard R list, `ggplot2` will apply them all when you add the list to a plot. This helps you prepare reusable components for your publication.

```{r}
#| label: demo-school-types
# Store reusable colour and shape scales for a few types of schools
school_types_scales <- list(
  scale_colour_manual(
    values = doe_colours("Primary School" = "blue-01",
                         "Central/Community School" = "red-02",
                         "Infants School" = "blue-02",
                         "Secondary School" = "grey-01")),
    scale_shape_manual(
      values = c("Primary School" = "square",
                 "Central/Community School" = "asterisk",
                 "Infants School" = "plus",
                 "Secondary School" = "circle open"))
)
```

Now we can use this scale in multiple plots, and have a consistent visual language.

```{r}
#| label: demo-multiple-metro-south
metro_south <- public_schools |>
  filter(
    Operational_directorate == "Metropolitan South",
    Level_of_schooling %in% c("Primary School",
                              "Central/Community School",
                              "Infants School",
                              "Secondary School"))

metro_south |>
  ggplot(aes(x = Date_1st_teacher,
             y = ICSEA_value,
             colour = Level_of_schooling,
             shape = Level_of_schooling)) +
  geom_point() +
  school_types_scales +
  theme_doe(base_size = 12)

metro_south |>
  ggplot(aes(x = ICSEA_value,
             y = FOEI_Value,
             colour = Level_of_schooling,
             shape = Level_of_schooling)) +
  geom_point() +
  school_types_scales +
  labs(x = "ICSEA", y = "FOEI") +
  theme_doe(base_size = 12)
```

The same logic applies to more than scales. If you want to bundle a lot of customisations, put them in a `list()` and add the list to your plot.

```{r}
#| label: demo-bundling
my_mega_customisation <- list(
  list(
  scale_colour_manual(
    values = doe_colours("Primary School" = "blue-01",
                         "Central/Community School" = "red-02",
                         "Infants School" = "blue-02",
                         "Secondary School" = "grey-01")),
    scale_shape_manual(
      values = c("Primary School" = "square",
                 "Central/Community School" = "asterisk",
                 "Infants School" = "plus",
                 "Secondary School" = "circle open")),
  theme_classic(paper = doe_colours("blue-04"))
  )
)

metro_south |>
  ggplot(aes(x= Date_1st_teacher,
             y = ICSEA_value,
             colour = Level_of_schooling,
             shape = Level_of_schooling)) +
  geom_point() +
  my_mega_customisation

metro_south |>
  ggplot(aes(x = ICSEA_value,
             y = FOEI_Value,
             colour = Level_of_schooling,
             shape = Level_of_schooling)) +
  geom_point() +
  labs(x = "ICSEA", y = "FOEI") +
  my_mega_customisation
```

## Make your own templates

If you need to make similar plots over and over, it's probably time to write a function. A simple pattern is to write a function that takes a dataframe and returns a `ggplot2` plot object.

```{r}
#| label: demo-function

# Bar plot: counts of schools, by type
plot_school_count <- function(df) {
  df |>
    # Order level of schooling by frequency
    mutate(school_type = fct_infreq(Level_of_schooling)) |>
    # Build our template plot
    ggplot(aes(y = school_type)) +
    geom_bar() +
    # Imrprove the appearance of the x axis
    scale_x_continuous(labels = scales::label_comma(),
                       breaks = scales::breaks_pretty()) +
    labs(x = "Number of schools",
         y = "Type of school") +
    theme_doe()
}
```

Because the function takes data and returns a `ggplot2`, you have flexibility to easily customise its inputs and further customise its output. Here's the base case, for all schools:

```{r}
#| label: demo-function-base-case
public_schools |>
  plot_school_count()
```

We can filter the input to a single principal network:

```{r}
#| label: demo-function-filter-input
public_schools |>
  filter(Principal_network == "Wagga Wagga") |>
  plot_school_count()
```

We can further customise the output:

```{r}
#| label: demo-function-customise-output
#| fig-height: 8
public_schools |>
  mutate(remoteness = fct_infreq(ASGS_remoteness)) |>
  plot_school_count() +
  facet_wrap(~remoteness, ncol = 1)
```

... and we can use our function with functionals and/or loops to really get going with automation:

```{r}
#| label: demo-functionals
public_schools |>
  filter(Operational_directorate == "Metropolitan South") |>
  group_by(Principal_network) |>
  group_map(~plot_school_count(.x) + labs(title = .y)) |>
  print()
```

### Exercise: factor out common parts of your workflow

Can you bundle several customisations so that you have a consistent, accessible and on-brand design language for these two plots?

```{r}
#| label: demo-factor-out
#| fig-height: 6
metro <- enrolled |>
  filter(grepl("Metro", Operational_directorate))

# Compare the relationship between ICSEA and LBOTE proportion,
# for each directorate, by school type
ggplot(metro,
       aes(x = ICSEA_value,
           y = LBOTE_pct,
           colour = Level_of_schooling)) +
  geom_point() +
  facet_wrap(~Operational_directorate,
             labeller = label_wrap_gen(20),
             ncol = 1)

# Compare the relationship between opening date and current ICSEA,
# by school type
ggplot(metro,
       aes(x = Date_1st_teacher,
           y = ICSEA_value,
           colour = Level_of_schooling)) +
  geom_point()
```

Make your changes here. 

```{r}
#| label: exercise-factor-common
#| exercise: factor-common
#| fig-height: 6

common_style <- list(
  ______
)

# Compare the relationship between ICSEA and LBOTE proportion,
# for each directorate, by school type
ggplot(metro,
       aes(x = ICSEA_value,
           y = LBOTE_pct,
           colour = Level_of_schooling,
           ______)) +
  geom_point() +
  facet_wrap(~Operational_directorate,
             labeller = label_wrap_gen(20),
             ncol = 1) +
  ______

# Compare the relationship between opening date and current ICSEA,
# by school type
ggplot(metro,
       aes(x = Date_1st_teacher,
           y = ICSEA_value,
           colour = Level_of_schooling,
           ______)) +
  geom_point() +
  ______
```

::: { .solution exercise="factor-common"}

```{r}
#| label: solution-factor-common
#| fig-height: 6
# Define common style elements as a list
common_style <- list(
  # Use common shape and colour scales
  scale_colour_manual(
    values = doe_colours("Primary School" = "blue-01",
                         "Secondary School" = "red-02")),
  scale_shape_manual(
    values = c("Primary School" = "circle",
               "Secondary School" = "plus")),
  # Use a consistent theme
  theme_doe(base_size = 12),
  theme(panel.border = element_rect(),
        legend.position = "right")
)

# Compare the relationship between ICSEA and LBOTE proportion,
# for each directorate, by school type
ggplot(metro,
       aes(x = ICSEA_value,
           y = LBOTE_pct,
           colour = Level_of_schooling,
           # Use shape as well as colour
           shape = Level_of_schooling)) +
  geom_point() +
  facet_wrap(~Operational_directorate,
             ncol = 1) +
  # Add our common styling
  common_style

# Compare the relationship between opening date and current ICSEA,
# by school type
ggplot(metro,
       aes(x = Date_1st_teacher,
           y = ICSEA_value,
           colour = Level_of_schooling,
           # Use shape as well as colour
           shape = Level_of_schooling)) +
  geom_point() +
  # Add our common styling
  common_style
```

:::

# Other packages

- [`colorspace`](https://cran.r-project.org/package=colorspace) provides more tools for assessing colours and palettes using contrast ratios and a variety of other measures. Includes tools for simulating colour blindness.
- [`ggpattern`](https://coolbutuseless.github.io/package/ggpattern/) helps you define and apply pattern fills. We didn't use any patterns today, but they are another way to avoid relying on colour alone.
- [`ggpackets`](https://dgkf.github.io/ggpackets/) provides a framework for easily templating `ggplot`s.
- [`ggrepel`](https://ggrepel.slowkow.com/) helps automate label placement to avoid overlapping text. Super useful if you have a lots of plots to make and don't want to hand-craft the label positions for each one.