---
title: "Automate on-brand, accessible data visualisations for your reporting using R and `doestyle`"
author: "Sam Gardiner (and you!)"
date: "4 June 2026"
format:
  html:
    fig-cap-location: margin
    reference-location: margin
    citation-location: margin
bibliography: references.bib
engine: knitr
knitr:
  opts_chunk: 
    dev: svglite
webr:
  packages:
    - doestyle
  repos:
    - https://nsw-education.r-universe.dev/
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

# Introduction 

This workshop has been developed and delivered on [Dharug](https://www.cityofparramatta.nsw.gov.au/community/first-nations/dharug-and-first-nations-community) country.

## Abstract

A picture tells a thousand words - but it can be time-consuming to craft data visualisations that meet our accessibility obligations while remaining on-brand and visually appealing. This workshop for R users will guide participants in the construction of several data visualisations using `doestyle`, the department’s in-house R package for plot styling. We will explore workflows to improve the ease and speed at which data visualisations can be built into evaluation reporting, and how to tweak these workflow defaults to improve visual appeal.

## Learning intentions

- Use R, `ggplot2` and `doestyle` to apply styling and theming to data visualisations
- Understand accessibility requirements for colour and contrast, and ensure that your data visualisations are accessible to the broadest section of our community
- Develop workflows to make building multiple plots for reporting as quick and simple as possible

# Getting started

This workshop is designed to be completed in a web browser. It uses [WebR](https://docs.r-wasm.org/webr/latest/) and [Quarto Live](https://r-wasm.github.io/quarto-live/) to run a [WebAssembly](https://webassembly.org/)-powered R session within your browser, with no installation required.

We'll be using a few packages throughout the rest of the workshop, so load them now:

```{r}
library(doestyle) #<1>
library(dplyr) #<2>
library(ggplot2) #<3>
```

As you loaded this web page, the `doestyle` R package was installed for your use during the workshop. However, if you want to use `doestyle` for your own work in RStudio or Visual Studio Code, you'll need to install it on your own computer:

```r
install.packages('doestyle', repos = c('https://nsw-education.r-universe.dev',
                                       'https://cloud.r-project.org'))
```

The `doestyle` package includes a dataset called `public_schools`, which is a snapshot of the [NSW public schools master dataset](https://www.metadata.nsw.gov.au/item/9544/distribution/nsw-public-schools-master-dataset).


Let's check that everything is working by building a quick visualisation with `ggplot2`, `doestyle` and the public schools data. Run the code and have a look at the output:

```{r}
#| label: fig-intro
#| fig-cap: Our first figure. Did it work? Is it accessible? Does it tell a story?

ggplot(public_schools,
       aes(y = Operational_directorate)) +
  geom_bar() +
  theme_doe() 
  #theme(geom = element_geom(paper = doe_colours("blue-01")))
```
# A quick tour of `ggplot2`, as helped by `doestyle`

## The Grammar of Graphics + `doestyle`

- link to Hadley's `ggplot2` paper
- link to Jonathan's talk
- `ggplot2` lets you map your data onto geometries via aesthetics and scales, building up plots in layers
- `doestyle` provides helpers for `ggplot2`:
  - Department of Education-branded scales
  - easy access to NSW Government-branded colours
  - a couple of custom geometries
  - tools to check colour contrast ratios and relative luminosity
  - tools to make it slightlier easier to use the Public Sans typeface
  
## Warm-up exercises: colours and themes

We'll use the copy of `public_schools` included within `doestyle`, and filter it so we're just looking at primary and secondary schools that have enrolments (`latest_year_enrolment_FTE`), a reported LBOTE percentage (`LBOTE_pct`) and a reported ICSEA (`ICSEA_value`).

```{r}
enrolled <- public_schools |>
  filter(Level_of_schooling %in% c("Primary School", "Secondary School")) |> #<1>
  filter_out( #<2>
    when_any(
      is.na(latest_year_enrolment_FTE),
      is.na(LBOTE_pct),
      is.na(ICSEA_value)
    )
  ) |>
  select(School_code, #<3>
         School_name,
         latest_year_enrolment_FTE,
         Level_of_schooling,
         LBOTE_pct,
         ICSEA_value)

# Take a peek:
glimpse(enrolled) #<4>
```

Let's visualise the relationship between school size and LBOTE percentage:

```{r}
ggplot(
  data = enrolled, #<1>
  mapping = aes( #<2>
    x = latest_year_enrolment_FTE, #<3>
    y = LBOTE_pct, #<4>
    colour = Level_of_schooling)) + #<5> 
  geom_point() + #<3>
  scale_x_continuous(labels = scales::label_comma()) #<4>
```

The plot works and tells a story, but it doesn't use the NSW Education brand palette. You can look up our colours in our brand guidelines, but the point of `doestyle` is to reduce the friction involved in making on-brand graphics:

```{r}
doestyle::show_colours()
```

If you need them, we can also use the whole NSW brand palette:

```{r}
#| column: page
#| fig-width: 10
#| fig-height: 4
show_colours("nsw")
```
We can also get the hex code for any named brand colours, without having to look at the swatches:

```{r}
doe_colours("blue-01")
```

Try it for yourself. Pick any two DoE or NSW colour names and retrieve their hex values:

```{r}
#| exercise: pick_colours

# doe_colours("______", "______")
doe_colours("blue-01", "red-02")
```


Now, use those hex codes in the plot we built earlier. We've added on `scale_colour_manual()`, which lets you create a mapping from the `Level_of_schooling` data to the aesthetics used by `geom_point`. Add the hex codes:

```{r}
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  # scale_colour_manual(values = c("______", "______")) #<1>
  scale_colour_manual(values = c("#002664", "#D7153A"))
```

Copying and pasting hex codes isn't a great workflow, so let's just use `doe_colours()` inside `scale_colour_manual()`:

```{r}
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  # scale_colour_manual(values = doe_colours("______", "______")) 
  scale_colour_manual(values = doe_colours("blue-01", "red-02"))
```

So that you don't have to look up colours for every plot, `doestyle` includes a ready-made scale and palettes that you can use instead of `ggplot2::scale_colour_manual()`: `scale_colour_doe()`.

```{r}
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_colour_doe() #<1>
```

Our `scale_colour_doe()` function has a `palette` argument. Its default is `doe_palettes$default`, which provides on-brand defaults for up to 12 categories^[Not that you should ever be trying to produce a plot that uses 12 colours - if so, it's time to rethink what you're trying to visualise!]:

```{r}
#| fig-height: 2
plot(doe_palettes$default)
```

Other built-in palettes are available in the `doe_palettes` object, which you can explore on your own.

If you want to define your own palettes, you can:

```{r}
my_custom_palette <- doe_colours("grey-01", "blue-02")

ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_colour_doe(palette = my_custom_palette) 
```

Finally, if you don't want to worry about setting scales or designing a palette, you can just apply `theme_doe()`, which will do its best to use sensible^[Sensible for a couple of colours, but after that you are better off doing your own decision-making.] defaults. You'll notice that `theme_doe()` applies some other defaults, liking removing the panel background, using Public Sans, increasing most text sizes and reducing the number of gridlines:

```{r}
ggplot(
  data = enrolled,
  mapping = aes(
    x = latest_year_enrolment_FTE,
    y = LBOTE_pct,
    colour = Level_of_schooling)) +
  geom_point() +
  theme_doe()
```

There's still plenty more we could do to polish this plot, but for now we'll move on to our main topic.

# Accessibility

From our very own website [@doeAccessibilityGuidelines2026]:

> Accessibility means providing equal access to information for all our audiences, with special consideration for people with disability.
> 
> ...
>
> Our approach to accessibility at the NSW Department of Education is about creating content that does not exclude anyone because of their ability, situation or circumstance. Rather, it is an inclusive approach to content that should make it easier for all users to understand, interact with and respond to our websites, apps, communications and materials.

- Accessibility improves the experience for everyone, not just people with a disability
- Starting your visualisations with accessibility in mind will often help you to design a simpler, more interpretable plot

## The standard: WCAG 2.2

We aim for level AA conformance with Web Content Accessibility Guidelines 2.2 [@wcag22].

> Web Content Accessibility Guidelines (WCAG) explains how to make web content more accessible to people with disabilities. WCAG covers web sites, applications, and other digital content.
>
> There are three levels of conformance:
>
> - Level A is the minimum level.
> - Level AA includes all Level A and AA requirements. Many organizations strive to meet Level AA.
> - Level AAA includes all Level A, AA, and AAA requirements. 

[@w3cwebaccessibilityinitiativewaiWebContentAccessibility]


You can read [the whole guideline](https://www.w3.org/TR/WCAG22/), but for our purpose today, we're most interested in sections of [Guideline 1.4 Distinguishable](https://www.w3.org/TR/WCAG22/#distinguishable):

> Make it easier for users to see and hear content including separating foreground from background.

### Colour

> #### [Success Criterion 1.4.1 Use of Color](https://www.w3.org/TR/WCAG22/#use-of-color)
>
> Color is not used as the only visual means of conveying information ...

:::{.callout-caution}
About 1 in 20 of your audience will have some form of colour blindness. If you use colour as the only way of communicating, you risk misinterpretation or even complete failure of your visusalisation.
:::

### Contrast

> #### [Success Criterion 1.4.3 Contrast (Minimum)](https://www.w3.org/TR/WCAG22/#contrast-minimum)
> 
> The visual presentation of text and images of text has a contrast ratio of at least 4.5:1 ...
> 
> Large-scale text and images of large-scale text have a contrast ratio of at least 3:1.

:::{.aside}
_Large text_ means 18 points, or 14 points and bold.
:::

- [Quick reference with examples](https://www.w3.org/WAI/WCAG22/quickref/#contrast-minimum)

## How can we meet the standard?

Use a second aesthetic mapping (an addition to colour) to represent categories. Depending on your plot type and data, there are many workable options:

- Colour + shape
- Colour + pattern or line type
- Colour + text annotation

If you have more than about 3 categories, there's a good chance you'll be better off designing a simpler visual:

- Can you move your categories to a labelled axis?
- Can you use small multiples instead (e.g. `ggplot2::facet_wrap()` or `facet_grid()`)?
- Would a table be easier to interpret?

### Exercise: getting away from colour alone

For this exercise, we'll look at schools with an attached preschool:

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)

preschools <- public_schools |>
  filter(Preschool_ind == "Y") |>
  filter_out(when_any(is.na(latest_year_enrolment_FTE),
                      is.na(ICSEA_value))) |>
  select(School_code,
         Level_of_schooling,
         Operational_directorate,
         latest_year_enrolment_FTE,
         ICSEA_value)

glimpse(preschools)
```

```{r}
ggplot(preschools,
       aes(y = Operational_directorate,
           fill = Level_of_schooling)) +
  geom_bar(position = position_dodge(preserve = "single"),
           width = 2/3) +
  coord_cartesian(expand = c(left = FALSE)) +
  labs(x = "Attached preschools") +
  theme_doe(base_size = 12)
```

```{r}
ggplot(preschools,
       aes(y = Operational_directorate,
           fill = Level_of_schooling,
           # Map level of schooling to another aesthetic, linetype:
           linetype = Level_of_schooling)) +
  geom_bar(position = position_dodge(preserve = "single"),
           width = 2/3,
           # Use the same line colour for every category
           color = "black") +
  scale_linetype_manual(values = c("Primary School" = "solid",
                                   "Central/Community School" = "dotted",
                                   "Infants School" = "dashed")) +
  coord_cartesian(expand = c(left = FALSE)) +
  theme_doe(base_size = 12)
```

```{r}
ggplot(preschools,
       aes(y = Level_of_schooling)) +
  geom_bar() +
  facet_wrap(~Operational_directorate,
             labeller = label_wrap_gen()) +
  coord_cartesian(expand = c(left = FALSE)) +
  theme_doe(base_size = 12)
```

```{r}
ggplot(preschools,
       aes(x = latest_year_enrolment_FTE,
           y = ICSEA_value,
           colour = Level_of_schooling)) +
  geom_point(size = 2) +
  theme_doe()
```

```{r}
ggplot(preschools,
       aes(x = latest_year_enrolment_FTE,
           y = ICSEA_value,
           colour = Level_of_schooling,
           shape = Level_of_schooling)) +
  geom_point(size = 2) +
  scale_shape_manual(values = c("Primary School" = "square",
                                "Central/Community School" = "asterisk",
                                "Infants School" = "plus")) +
  theme_doe()
```

### Exercise: use high-contrast text annotations

Sometimes your audience will ask for data labels or other text annotations. 

```{r}
lbote_topn <- public_schools |>
  filter_out(is.na(LBOTE_pct)) |>
  slice_max(LBOTE_pct, n = 1, by = Operational_directorate, with_ties = FALSE) |>
  mutate(
    school_label = paste0(School_name, ": ", LBOTE_pct, "% "),
    school_label_fct = forcats::fct_reorder(school_label, LBOTE_pct))

ggplot(lbote_topn,
       aes(x = LBOTE_pct,
           y = school_label_fct)) +
  geom_col() +
  geom_text(aes(label = school_label),
            hjust = "right",
            size = 12 / .pt) +
  scale_x_continuous(labels = \(x) paste0(x, "%")) +
  theme_doe() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank()) +
  labs(x = "Students with Language Background Other Than English (LBOTE)",
       y = NULL)
```
`doestyle` provides some help with choosing contrasting colours for text:

```{r}
doestyle::show_colours()
```



```{r}
contrast_ratio(doe_colours("blue-01"), doe_colours("blue-04"))
```



### Department of Education guidance

- https://brand.education.nsw.gov.au/content/edam/en/brand-guideline.html

